Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data

### 2022-08-18 analysis/paper/paper.Rmd
moving captions to end

---
title: "Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data"
author:
  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^{a,b}$, ...
  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Potsdam, Germany*
  - $^b$ *Deutsches Archäologisches Institut, Dendroclimatology, Berlin, Germany*
  - "* Corresponding author: hurley@gfz-potsdam.de"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::html_document2:
      fig_caption: yes
      toc: false
      number_sections: false
      global_numbering: true
    bookdown::word_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
      global_numbering: true
    bookdown::pdf_document2:
      fig_caption: yes
      toc: yes
      number_sections: false
      global_numbering: true
header-includes:
- \usepackage{amsmath}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
bibliography: 
  - uhi_references.bib
  - 04_gfz_UHI.bib
  - 07_phenoTram.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The socio-ecological benefits that urban trees provide depend on their health and performance. Excess heat (i.e., Urban Heat Island; UHI) and other urban conditions affect tree physiology with outcomes from enhanced growth to mortality. Resilient urban forests in the face of climate change require species-specific understanding of growth responses. However, to date, studies assessing growth dynamics were primarily based on remote sensing of communities rather than individuals, or relied on labor-intensive methods that can limit the spatial coverage necessary to account for highly variable urban growing conditions. Here, we analyze growth dynamics of common urban tree species over time and across space for Berlin (Germany), a large metropolis, combining dendroecological (temporal) and inventory assessments (spatial). First, we show that annual increments increased across the 20th century for early (i.e., young) growth. Second, we use an approach relying on open inventory data to identify growth potential (diameter) in relation to excess heat across Berlin while accounting for age, potential management effects, and the urban fabric (e.g., planting area; building density, height; available soil nutrients) by applying generalized additive models for the ten most abundant species. Our analyses showed that younger trees may benefit from increased temperatures, while older individuals feature lower growth at greater UHI magnitudes. Furthermore, we show that planting area as well as building density modulate growth responses to temperature. Lastly, we discuss implications for the management of urban trees in the context of climate change mitigation, considering that younger trees are predominantly located at UHI “hot spots” and will undergo the observed age-dependent shift in temperature-growth sensitivity. By relying on increasingly available open data (inventories, urban/environmental data) our approach here is or will be readily transferable to other urban regions, allowing for species-specific growth assessments, while accounting for urban conditions at an individual level.

@@

(ref:study-area) Berlin's generalized land-use derived from @suvkberlin2019 and location within the European context (inset).

```{r fig-study-area, fig.cap="(ref:study-area)",fig.width = 12, fig.height = 14, out.width="50%"}

@@

@@

@@

(ref:temp-overview) Overview of data sets used for assessing the relationship between temperature and tree growth, with data sources in rows and averaged time intervals in columns. Note the varying spatial coverage and resolution as noted in Table \@ref(tab:tab-dsources).

```{r fig-temp-overview, fig.cap="(ref:temp-overview)",fig.width = 12, fig.height = 12, out.width="100%"}

@@

@@

(ref:cambial-effect) Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section [Dendrochronological analyses](#sec:methods-dendro). Thick lines and bands are for mean and 95$~\%$ confidence interval of all annual predictions across a time group (fine lines). On average, early stages of recent growth (up to approximately 22 years) exceed that of earlier periods discernibly.

```{r fig-cambial-effect, fig.cap="(ref:cambial-effect)", out.width="50%", include=TRUE, eval=TRUE}

@@

# Appendices
Information too detailed to be included in the main text, for instance a list of areas sampled, may be presented in appendices which are to be placed before the References. Appendices should be as short and succinct as possible.

@@

<!-- ### Colophon -->

<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies:  -->

<!-- ```{r colophon, cache = FALSE} -->
<!-- # which R packages and versions? -->
<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
<!-- ``` -->

<!-- The current Git commit details are: -->

<!-- ```{r} -->
<!-- # what commit is this file at?  -->
<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
<!-- ``` -->

>>>>

---
title: "Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data"
author:
  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^{a,b}$, ...
  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Potsdam, Germany*
  - $^b$ *Deutsches Archäologisches Institut, Dendroclimatology, Berlin, Germany*
  - "* Corresponding author: hurley@gfz-potsdam.de"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::html_document2:
      fig_caption: yes
      toc: false
      number_sections: false
      global_numbering: true
    bookdown::word_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
      global_numbering: true
    bookdown::pdf_document2:
      fig_caption: yes
      toc: yes
      number_sections: false
      global_numbering: true
header-includes:
- \usepackage{amsmath}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}

bibliography: 
  - uhi_references.bib
  - 04_gfz_UHI.bib
  - 07_phenoTram.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The socio-ecological benefits that urban trees provide depend on their health and performance. Excess heat (i.e., Urban Heat Island; UHI) and other urban conditions affect tree physiology with outcomes from enhanced growth to mortality. Resilient urban forests in the face of climate change require species-specific understanding of growth responses. However, to date, studies assessing growth dynamics were primarily based on remote sensing of communities rather than individuals, or relied on labor-intensive methods that can limit the spatial coverage necessary to account for highly variable urban growing conditions. Here, we analyze growth dynamics of common urban tree species over time and across space for Berlin (Germany), a large metropolis, combining dendroecological (temporal) and inventory assessments (spatial). First, we show that annual increments increased across the 20th century for early (i.e., young) growth. Second, we use an approach relying on open inventory data to identify growth potential (diameter) in relation to excess heat across Berlin while accounting for age, potential management effects, and the urban fabric (e.g., planting area; building density, height; available soil nutrients) by applying generalized additive models for the ten most abundant species. Our analyses showed that younger trees may benefit from increased temperatures, while older individuals feature lower growth at greater UHI magnitudes. Furthermore, we show that planting area as well as building density modulate growth responses to temperature. Lastly, we discuss implications for the management of urban trees in the context of climate change mitigation, considering that younger trees are predominantly located at UHI “hot spots” and will undergo the observed age-dependent shift in temperature-growth sensitivity. By relying on increasingly available open data (inventories, urban/environmental data) our approach here is or will be readily transferable to other urban regions, allowing for species-specific growth assessments, while accounting for urban conditions at an individual level.

@@

**Fig. \@ref(fig:fig-study_area)**

@@

**Table Table$~$\@ref(tab:tab-dsources)**

@@

**Table \@ref(tab:tab-age)**

@@

**Fig. \@ref(fig:fig-temp-overview)

@@

**Table \@ref(tab:tab-dendro-sites)**

@@

**Figure \@ref(fig:fig-cambial-effect)**

@@

# Appendix {-}

@@

# Figures

## Legends

**Figure \@ref(fig:fig-study-area)** Berlin's generalized land-use derived from @suvkberlin2019 and location within the European context (inset).

**Figure \@ref(fig:fig-temp-overview)** Overview of data sets used for assessing the relationship between temperature and tree growth, with data sources in rows and averaged time intervals in columns. Note the varying spatial coverage and resolution as noted in Table \@ref(tab:tab-dsources).

**Figure \@ref(fig:fig-cambial-effect)** Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section [Dendrochronological analyses](#sec:methods-dendro). Thick lines and bands are for mean and 95$~\%$ confidence interval of all annual predictions across a time group (fine lines). On average, early stages of recent growth (up to approximately 22 years) exceed that of earlier periods discernibly.

## Visualization

```{r fig-study-area, fig.cap = "PLACEHOLDER", fig.width = 12, fig.height = 14, out.width="50%"}

```{r fig-temp-overview, fig.cap="PLACEHOLDER",fig.width = 12, fig.height = 12, out.width="100%"}

```{r fig-cambial-effect, fig.cap="PLACEHOLDER", out.width="50%", include=TRUE, eval=TRUE}

# Tables
